---
title: "HW4"
author: "Farooq Mahmud"
format: pdf
editor: source
---

## Problem 3a


```{r}
library(TSA)
data(electricity)
acf(electricity)
```
The data does not appear to be stationary based on the ACF plot because:

- Significant ACF at all lags.
- Seasonal pattern with peaks every 12 lags.
- Slow, if any decay in ACF.

## Problem 3b
```{r}
log_electricity <- log(electricity)
diff_log_electricity <- diff(log_electricity)
acf(diff_log_electricity)
```

Yes, seasonality is visible. Since the electricity data is monthly and there are significant ACF's at every 12 lags, the seasonal component is 12 months.

## Problem 3c
```{r}
diff_seasonal <- diff(diff_log_electricity, lag = 12)
plot(diff_seasonal)
```
Yes, the plot looks stationary now.

## Problem 3d
```{r}
acf(diff_seasonal)
```
- $p=0$ because no gradual decay appears in ACF.
- $d=1$ because we needed first differencing to remove trend/non-stationarity.
- $q=1$ because of a sharp drop in ACF after lag 1.
- $P=0$ because no gradual decay or spike pattern.
- $D=1$ because seasonal differencing of order 1 is needed.
- $Q=1$ because the large spike at lag 12.
- Seasonal period of 12

Therefore a model to consider would be:

$ARIMA(0,1,1) X (0,1,1)_{12}$

## Problem 4a

```{r}
library(TSA)
data("airmiles")
acf(airmiles)
pacf(airmiles)
```
Just going by the ACF and PACF:

- $p=0$ because there is no sharp cutoff in PACF.
- $d=1$ because the series appears non-stationary and a first differencing is likely needed.
- $q=1$ because ACF cuts of after lag 1 which suggests $MA(1)$.
- $P=0$ because PACF does not show a seasonal spike at lag 12 suggesting no strong seasonal AR.
- $D=1$ because ACF shows spikes every 12 lags suggesting one seasonal difference is needed.
- $Q=1$ because of the significant spike at lag 12 suggesting seasonal $MA(1)$.
- $s=12$ because this is monthly data.

# Problem 4b

```{r}
fit_b <- arima(airmiles, 
               order = c(0, 1, 1), 
               seasonal = list(order = c(0, 1, 1), period = 12))

ts.plot(residuals(fit_b), main = "Residual Plot", ylab = "Residuals")
acf(residuals(fit_b), main = "ACF of Residuals")

```
$ARIMA(0,1,1) X (0,1,0)_{12}$ appears to be a reasonable model because:

- The residual plot shows no obvious trend or seasonality.
- The ACF of residuals has all non-significant values.

# Problem 4c

```{r}
library(forecast)
fit_auto <- auto.arima(airmiles)
summary(fit_auto)
```
# Problem 4d

```{r}
model_d <- arima(airmiles, 
                 order = c(1, 0, 1), 
                 seasonal = list(order = c(0, 1, 1), period = 12))

ts.plot(residuals(model_d), main = "Residual Plot", ylab = "Residuals")
acf(residuals(model_d), main = "ACF of Residuals")
```
$ARIMA(1,0,1) X (0,1,1)_{12}$ suggested by `auto.arima` appears to be a reasonable model because:

- The residual plot shows no obvious trend or seasonality.
- The ACF of residuals has all non-significant values.