---
title: "HW4"
author: "Farooq Mahmud"
format: pdf
editor: source
---

## Problem 3a


```{r}
library(TSA)
data(electricity)
acf(electricity)
```
The data does not appear to be stationary based on the ACF plot because:

- Significant ACF at all lags.
- Seasonal pattern with peaks every 12 lags.
- Slow, if any decay in ACF.

## Problem 3b
```{r}
log_electricity <- log(electricity)
diff_log_electricity <- diff(log_electricity)
acf(diff_log_electricity)
```

Yes, seasonality is visible. Since the electricity data is monthly and there are significant ACF's at every 12 lags, the seasonal component is 12 months.

## Problem 3c
```{r}
diff_seasonal <- diff(diff_log_electricity, lag = 12)
plot(diff_seasonal)
```
Yes, the plot looks stationary now.

## Problem 3d
```{r}
acf(diff_seasonal)
```
- $p=0$ because no gradual decay appears in ACF.
- $d=1$ because we needed first differencing to remove trend/non-stationarity.
- $q=1$ because of a sharp drop in ACF after lag 1.
- $P=0$ because no gradual decay or spike pattern.
- $D=1$ because seasonal differencing of order 1 is needed.
- $Q=1$ because the large spike at lag 12.
- Seasonal period of 12

Therefore a model to consider would be:

$ARIMA(0,1,1) X (0,1,1)_{12}$



