---
title: "HW3"
author: "Farooq Mahmud"
format: pdf
editor: source
---

## Problem 5a


```{r}
phi <- c(0.7, -0.4)

theoretical_acf <- ARMAacf(ar = phi,
                           ma = numeric(0),
                           lag.max = 20)
plot(
  theoretical_acf,
  type = "h",
  main = "Theoretical ACF",
  xlab = "Lag",
  ylab = "ACF"
)
abline(h = 0)
```
## Problem 5b
```{r}
set.seed(123)

phi <- c(0.7, -0.4)
n <- 72
ar2_series <- arima.sim(n = n, model = list(ar = phi))

acf(ar2_series, main = "Sample ACF")

```

The sample ACF closely resembles the theoretical ACF in shape. The overall pattern of slow decay after significant values in lags 1 and 2 shows up in both plots which is congruent with this being an AR(2) time series.

## Problem 5c
```{r}
theoretical_pacf <- ARMAacf(
  ar = phi,
  ma = numeric(0),
  lag.max = 20,
  pacf = TRUE
)

plot(
  theoretical_pacf,
  type = "h",
  main = "Theoretical PACF",
  xlab = "Lag",
  ylab = "PACF"
)
abline(h = 0)
```

## Problem 5d
```{r}
pacf(ar2_series, main = "Sample PACF")
```
The sample PACF matches well with the theoretical ACF. There are significant spikes at lags 1 and 2 and insignificant values after. This is congruent with this being an AR(2) time series.

